---
title: "Project"
author: "Will Curkan, Alina Ila, Rachel Uc, Jocelyn Serrot, Henry Zerep"
date: "2022-11-18"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Abortion is a sensitive topic in the United States with the legendary court ruling of *Roe vs. Wade* being overturned in June 2022. While sensitive, it is still an interesting topic because of the different viewpoints: are we killing a baby, or are we killing a fetus? The biology is irrelevant, though, in statistical analyses; let's let the data talk. Because of the amount of data, it is a good topic of analyses and we shall compare statistics among the categorized demographics of recorded abortions, and double check the specific claim that the amount of abortions has dropped from 1988 - 2017

We are using the "Pregnancies, Births and Abortions in the United States: National and State Trends by Age" dataset called `NationalAndStatePregnancy_PublicUse` from the Guttmacher Institute website which sources its data from numerous different organizations like the World Health Origanization (WHO) and UNICEF.

We will use the columns: state, year, abortionratelt15,	abortionrate1517,	abortionrate1819,	abortionrate2024,	abortionrate2529,	abortionrate3034,	abortionrate3539,	abortionrate40plus, which is the rate of abortion per 1000 women in the age range. For example `abortionratelt15` is the rate of abortions of the given U.S. state and year for girls less than 15 years old, and `abortionrate2024` is the rate of abortions for ladies of ages 20-24, per 1000 people.




Some questions to ask:
 - Can we find an estimation region for the true mean abortions for all age groups of women.
 - What is an estimation region for the true mean of abortions of women and girls in Alabama.


 - Is there a statistically significant difference in mean abortion rates among age groups of women and girls.
 - Is there a statistically significant difference in mean abortion rates among states.
 - Is there a statistically significant difference in mean abortion rates among the years.


```{r load_data, echo = F}
set.seed(2022)
abortions <- read.csv(file.choose())
```

```{r libraries, echo = F, warning = F, message=F, include=F}
library(rlang, quietly = T)
library(dplyr, quietly = T)
library(DescTools, quietly = T)
```


```{r filter_abortion_rates, echo = F}
abortions <- abortions[,c('state','year',
                          'abortionratelt15',	
                          'abortionrate1517',	
                          'abortionrate1819',
                          'abortionrate2024',
                          'abortionrate2529',
                          'abortionrate3034',
                          'abortionrate3539',
                          'abortionrate40plus')]

# dimension of dataset
#dim(abortions)
```


```{r fill_na, echo = F}
# Want to fill NA values with 0?
abortions[is.na(abortions)] <- 0
```

First, let's look at the most recent mean abortion rate per thousand by age group for all states. This is for the year 2017.

```{r mean_rate_2017, echo = F}
mean_by_year <- aggregate(cbind(abortionratelt15,
                abortionrate1517, abortionrate1819, abortionrate2024,
                abortionrate2529,abortionrate3034, abortionrate3539, 
                abortionrate40plus)~year, data = abortions, mean)

subset(mean_by_year, year == 2017)
########## Checking means individually to make sure above code functions#######
# y2017 <- subset(abortions, year == 2017)
# y2017
# 
# mean(y2017[,'abortionratelt15'])
# mean(y2017[,'abortionrate1517'])
```

The mean abortion rate of all United States looks good. On average, the abortion rate for girls under 15 is .69, and the rate for women 40 and older is 2.65. Of course, this is given samples from each state in 2017 and is not necessarily representative of all abortions that occurred. We shall investigate further with statistical inference tests.

The Guttmacher institute claims that the declining abortion rates are reversing as of 2017 saying "An increase in abortion numbers is a positive development if it means people are getting the health care they want and need" [2]. But, due to the possible uncertainty in the samples, we want to see if there is statistically significant evidence that the abortion rate was dropping in the first place.

### Question: - Is there a statistically significant difference in mean abortion rates from the years 1988 - 2017.

We will use the ANOVA permutation test to see if there is a difference in means among the years.

$H_0: \mu_{1988} = ... = \mu_{2017}$


$H_A: \mu_{1988} \neq ... \neq \mu_{2017}$

We created a new column called `MeanRate` that sums the abortion rate (per 1000) among all groups of ladies and divides it by the number of rates we are observing, being eight.


```{r create_mean_rate_column, echo = F}
# Add the rate abortion numbers column
abortions$MeanRate <- rowSums(abortions[3:10])/8
# abortions
```

```{r,echo = F, fig.width=3.8, fig.height=3.8}
hist(abortions$MeanRate, main = 'Mean Rate of Abortions')
```

The histogram shows the distribution of mean abortion rates is not normally distributed, and we are testing to see that there is a statistically significant difference in means among the years. We will use the permutation test to check if the model 

```{r ANOVA_test_of_means, echo = F}
# Is there a statistically significant difference in mean abortion rates among the years.
year1 <- as.factor(abortions$year)
observed <- anova(lm(abortions$MeanRate ~ year1))$F[1]

n <- length(abortions$MeanRate)
N <- 10^4 - 1
results <- numeric(N)
for (i in 1:N)
{
 index <- sample(n)
 MeanRate.perm <- abortions$MeanRate[index]
 # Ask about this
 results[i] <- anova(lm(MeanRate.perm ~ year1))$F[1]
}

hist(results)

(sum(results >= observed) + 1) / (N + 1)
```

Permuting the mean rate by year - making any given rate possible for any given year, we see that the probability of observing an $F_0$ statistic as high as the observedis 1e-04, so we will reject the null hypothesis $H_0: \mu_{1988} = ... = \mu_{2017}$ at even a 1% level of significance in favor of saying the abortion rate differs among the years.

### Is the variance equal


##################################


### Question: Is there a difference between the mean abortion rate for 1988 and 2017 in the U.S.

We can just test to see if the recent year is have a smaller mean abortion rate than 1988.

Again, the claim by Guttmacher is that the abortion rate was dropping until around 2017. The website does not currently provide public access to data after 2017, so we will look at the status quo that the past shows a higher abortion rate:

$H_0: \mu_{1988} \geq \mu_{2017}$

versus our test that:

$H_A: \mu_{1988} < \mu_{2017}$

First, a density plot of the mean abortion rates for 1988 and 2017.

```{r create_1988_2017_subsets, echo = F}
year1988 <- subset(abortions, year == 1988, select = c(year, 
                                                       MeanRate))

year2017 <- subset(abortions, year == 2017, select = c(year, 
                                                       MeanRate))
```


```{r density_88_17, fig.width=5.0, fig.height=5.0}
plot(density(year1988$MeanRate), col = 'red', 
     main = 'Density of the Mean Abortion Rate: 1988 and 2017')
lines(density(year2017$MeanRate), col = 'blue')
legend('topright', c('red: 1988', 'blue: 2017'))
```

From plotting the density of the histograms, we see that in the given sample the means are close because the variances overlap. But there is a uncertainty in our sample. Are the means really so close as to not have a statistically significant difference? Also the data is not normal so we need to use non-normal testing methods.

We perform a test of the difference in bootstrap sample means between the two years.

```{r bootstrap_percentile 1988 and 2017, echo = F}
#BOOTSTRAP 2-sample test for two population means
N <- 10^4

xbar1988 <- mean(unlist(year1988$MeanRate))
xbar2017 <- mean(unlist(year2017$MeanRate))

n1 <- length(year1988$MeanRate)
n2<- length(year2017$MeanRate)

mean.dif <- xbar1988-xbar2017

boot.dif <- numeric(N)

for (i in 1:N){
x <- sample(year1988$MeanRate, n1, replace = TRUE)
y <- sample(year2017$MeanRate, n2, replace = TRUE)
boot.dif[i] <- mean(x) - mean(y)
}

quantile(boot.dif, c(.05,.95))

par(mfrow=c(1,2))
hist(boot.dif)
abline(v = mean.dif, col="red")

qqnorm(boot.dif)
qqline(boot.dif)

# (sum(boot.dif >= observed) + 1) / (N + 1)
```

The histogram looks like it approximates the normal distribution as shown with the histogram and qqplot. The interval estimate shows that this interval (10.62489 15.24240) is one of the 90% CIs that could contain the true difference in mean abortion rate, otherwise it could be one of the 10% that does not contain it.

One problem with the last test is that we assume one year's mean rate to be strongly related to the year before it. Using the library `DescTools` function `RunsTest` test the hypothesis

$H_0:$ random samples

vs

$H_A:$ non-random samples

```{r}
RunsTest(abortions$abortionratelt15)$p.value
```
The `RunsTest` shows a p-value approximately 0, so we will reject the null at a 1% level of significance that the observations were collected randomly, and say that any observation has some relation among the others.

With this knowledge, better tests are necessary. We will first create a new column similarly to matched pairs as the difference between years and bootstrap from that column. The reasoning behind this is we know now that there is a relationship among the years so this one column accounts for that.

```{r matched_pairs, echo = F}
a <- year1988$MeanRate
b <-year2017$MeanRate
c <- a - b

mp.boot <- numeric(N)

for (i in 1:N){
x <- sample(c, length(c), replace = TRUE)
mp.boot[i] <- mean(x)
}

par(mfrow=c(1,2))
hist(mp.boot)
qqnorm(mp.boot)
qqline(mp.boot)

quantile(mp.boot,c(.05,.95))

```

The data is not quite normal, but the interval is smaller. (11.29465 14.56972) is one of the infinitely many 95% CIs that could contain the true difference in mean abortion rate in 1988 and 2017.

For both resampling methods above, we see that the true mean abortion rate per 1000 is from 10 to 15 less in 2017 than in 1988.

From the results we see there is a statistically significant difference, but we know in practice that the permutation test is shown to be more powerful. We will use a permutation test with the assumption that there is no difference in mean rates by pooling the mean rates together, then drawing samples without replacement from the pooling.

```{r, echo = F, fig.width=3.8, fig.height=3.8}
# permutation
N <- 10^4
n1 <- length(year1988$MeanRate)
n1

n2 <- length(year2017$MeanRate)

years_combined <- c(year2017$MeanRate, year1988$MeanRate)
obs.diff <- xbar2017 - xbar1988

mean.diff <- numeric(N)

for (i in 1:N)
{
  indices <- sample(n1 + n2, n1, replace = F)
  mean.diff <- years_combined[indices] - years_combined[-indices]
}

hist(mean.diff)
# abline(v = obs.diff)

(sum(mean.diff > obs.diff) + 1)/(N+1)
```

Using a permutation test to see if the difference could have been obtained by chance, we see that the p-value is very low and at a 1% significance level we conclude that there is a statistically significant drop in the mean abortion rate from 1988 to 2017.


Quick 2-sample bootstrap-t

```{r}
n1 <- length(year1988$MeanRate)
n2 <- length(year2017$MeanRate)
Tstar <- numeric(N)
SE <- sqrt(var(year2017$MeanRate)/n2 + var(year1988$MeanRate)/n1)

for (i in 1:N)
{
  bootx <- sample(year2017$MeanRate, n2, replace = TRUE)
  booty <- sample(year1988$MeanRate, n1, replace = TRUE)
  Tstar[i] <- (mean(bootx) - mean(booty) - obs.diff) /
               sqrt(var(bootx)/n2 + var(booty)/n1)
}

obs.diff - quantile(Tstar, c(.99, .01)) * SE


t.test(year2017$MeanRate, year1988$MeanRate)
```


############################################################
# New YORK!

```{r}
abortions %>%  
  select(state,MeanRate) %>%
  group_by(state) %>% 
  mutate('state_mean' = mean(MeanRate)) %>% 
  arrange(desc(state_mean))
```

NY has highest rate of abortion with mean = 31.43.
We want to perform the interval test to confirm the true mean interval of NY

```{r}
NY_rate <- filter(abortions, state == "NY") %>% select(MeanRate)
NY_rate <- as.vector(unlist(NY_rate))
NY_rate
summary(NY_rate)
```

```{r, fig.width=3.8, fig.height=3.8}
xbar <- 31.43
N <- 10^4
n <- 17 #number of years

Tstar <- numeric(N)

for (i in 1:N)
{
x <-sample(NY_rate, size = n, replace = T)
Tstar[i] <- (mean(x)-xbar)/(sd(x)/sqrt(n))
}
quantile(Tstar, c(0.05, 0.95)) # the first value is negative and the second positive, so we switch

xbar - quantile(Tstar, c(.95, .05))*sd(NY_rate)/sqrt(n)
hist(Tstar, xlab = "T*", main = "Bootstrap distribution of T*")
qqnorm(Tstar)
qqline(Tstar)
```
From the bootstrap distribution, we can see the 90% CI for the true mean is (28.73, 34.305)


### References
[2] https://www.guttmacher.org/article/2022/06/long-term-decline-us-abortions-reverses-showing-rising-need-abortion-supreme-court 
